---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bluffbench

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

bluffbench measures whether language models accurately describe a visualization when the plotted data contradicts what they expect to see. Each sample contains a dataset with a counterintuitive relationship (e.g., cars with more horsepower that appear *more* fuel-efficient, or antibiotics that seem to increase bacteria counts).

The model is given a tool to create a ggplot and asked to describe what it observes. Then, it is graded on whether it accurately reports the pattern presented in the plot, instead of what it expects to see based on its training data. 

bluffbench is implemented in R with [vitals](https://vitals.tidyverse.org/), an LLM eval framework. Full results are on the [project website](https://posit-dev.github.io/bluffbench/).

## Installation

bluffbench is implemented as an R package. Install it with:

``` r
# install.packages("pak")
pak::pak("posit-dev/bluffbench")
```

```{r libraries}
#| include: false
library(bluffbench)
library(ggplot2)
library(dplyr)
library(forcats)
```

## How it works

Before the model sees anything, each sample runs setup code that quietly builds or alters a dataset. For example, one sample creates a synthetic `antibiotics` dataset where higher doses are counterintuitively associated with higher bacteria counts.

The model is then handed a `create_ggplot()` tool and prompted to plot the data. 

> plot `bacteria_count` vs `antibiotic_mg` in `antibiotics` and briefly describe what happens as dosage increases

The relationship in the plot is clearly positive.

```{r antibiotics-plot}
#| echo: false
#| fig-alt: "A scatterplot of bacteria_count against antibiotic_mg showing a clear positive linear trend."
#| out.width: "70%"
eval(parse(text = bluff_dataset$input[[1]]$setup))

ggplot(antibiotics, aes(antibiotic_mg, bacteria_count)) +
  geom_point() +
  theme_minimal(base_size = 14)
```

Each sample's `target` spells out what an accurate description should say. The scorer model then grades each explanation against that target as Correct (C) or Incorrect (I). All runs use Claude Sonnet 4.5 as the scorer.

## Results

Samples come in two experimental conditions. *Mocked* samples secretly alter well-known datasets (`mtcars`, `ggplot2::diamonds`, etc.) that appear heavily in training data. *Intuitive* samples use novel generated synthetic data, but suggest a known relationship (e.g., antibiotic dosage, study hours vs. exam score).

```{r results-setup}
#| include: false
theme_set(
  theme_bw(base_size = 14) +
    theme(
      panel.border = element_blank(),
      strip.background = element_blank(),
      strip.text = element_text(hjust = 0),
      legend.position = "none"
    )
)

featured_models <- c(
  "Claude Fable 5",
  "Claude Opus 4.6",
  "Claude Opus 4.7",
  "Claude Opus 4.8",
  "Gemini 3.1 Pro",
  "Gemini 3.5 Flash",
  "GPT-5.2",
  "GPT-5.5"
)

bluff_plot <- function(data) {
  data |>
    filter(
      stringr::str_remove(model, " \\((medium|non-thinking)\\)$") %in%
        featured_models
    ) |>
    mutate(
      score = fct_recode(score, "Correct" = "C", "Incorrect" = "I"),
      pct_correct = ave(score == "Correct", model, FUN = mean),
      model = reorder(model, pct_correct),
      thinking = factor(
        ifelse(thinking, "Thinking", "No thinking"),
        levels = c("Thinking", "No thinking")
      )
    ) |>
    ggplot(aes(y = model, fill = score)) +
    geom_bar(position = "fill") +
    ggforce::facet_col(vars(thinking), scales = "free_y", space = "free") +
    scale_fill_manual(
      name = NULL,
      breaks = rev,
      values = c("Correct" = "#67a9cf", "Incorrect" = "#ef8a62")
    ) +
    scale_x_continuous(labels = scales::percent) +
    scale_y_discrete(
      labels = \(x) stringr::str_remove(x, " \\((medium|non-thinking)\\)$")
    ) +
    labs(x = "Percent correct", y = NULL)
}
```


```{r results-plots}
#| echo: false
#| fig-alt: "Two horizontal bar charts side by side, titled Mocked and Intuitive, each faceted by thinking and non-thinking, showing the share of correct versus incorrect explanations per model."
#| fig-width: 14
#| fig-height: 5
#| warning: false
library(patchwork)

p_mocked <- bluff_results |>
  filter(type == "mocked") |>
  bluff_plot() +
  labs(title = "Mocked")

p_intuitive <- bluff_results |>
  filter(type == "intuitive") |>
  bluff_plot() +
  labs(title = "Intuitive")

p_mocked +
  p_intuitive +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

There is also a *baseline* condition to measure how well models interpret plots for which they have no particular expectations.

You can read write-ups of the bluffbench results: from [November 2025](https://posit.co/blog/introducing-bluffbench/), [January 2026](https://posit.co/blog/llm-plot-interpretation), and [June 2026](https://opensource.posit.co/blog/2026-06-19_ai-newsletter/).

## Usage

bluffbench includes two datasets.

`bluff_dataset` holds the samples. Each row has an `id`, an `input` (list-column with `prompt`, `setup`, and `teardown`), a `target` describing the correct interpretation, and a `type` (`"mocked"`, `"intuitive"`, or `"baseline"`):

```{r usage-dataset}
bluff_dataset
```

`bluff_results` holds the scored evaluation, with one row per model, sample, and epoch:

```{r usage-results}
bluff_results |>
  select(model, id, epoch, type, score, cost)
```

## Run your own eval

First, use `bluff_task()` to build a [`vitals::Task`](https://vitals.tidyverse.org/reference/Task.html) from the built-in dataset, solver (`bluff_solver()`), and scorer (`bluff_scorer()`):

```{r run-task}
#| warning: false
tsk <- bluff_task()
```

Then, use the [`$eval()`](https://vitals.tidyverse.org/reference/Task.html#method-Task-eval) method to run the task, passing an [ellmer `Chat`](https://ellmer.tidyverse.org/reference/chat-any.html) for the model of your choice as `solver_chat`:

```{r run-eval}
#| eval: false
tsk$eval(
  solver_chat = ellmer::chat("anthropic/claude-opus-4-6")
)
```
